Natural Selection Across Three Generations of Americans

### 2024-02-23 hrs-selection.qmd
moved discussion of non-result to discussion section

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  and unmarried parents, but not among younger parents or those with less 
  education. Using respondents' number of siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
  natural selection: scores which correlate negatively (positively) with
  partly borne out: coefficients are larger among low-income
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

##############################################################
# Residualising then standardising PGS, with sampling weights
##############################################################

dat_eth_list <- list()

  r_eth <- rand |> subset(ethnicity == eth)

for (pg in pgs) {

  reg <- lm(get(pg) ~ PC1_5A + PC1_5B + PC1_5C + PC1_5D + PC1_5E +
                    PC6_10A + PC6_10B + PC6_10C + PC6_10D +PC6_10E, weights = weight, 
                    data = r_eth)

  r_eth <- r_eth |> mutate(
                    !!pg := get(pg) - predict(reg, r_eth),
                    mean_weight = mean(weight, na.rm=TRUE),
                    tot_weight = sum(weight, na.rm=TRUE),
                    mean_res_pgs = sum(get(pg)*weight, na.rm=TRUE)/tot_weight,
                    w.dev = weight*(get(pg) - mean_res_pgs)^2,
                    w.var.pgs = sum(w.dev, na.rm =TRUE)/tot_weight,
                    deman_pgs = get(pg) - mean_res_pgs,
                    !!pg := deman_pgs/sqrt(w.var.pgs)
  )
}

  dat_eth_list[[eth]] <- r_eth

}

rand <- dat_eth_list |> list_rbind()
pgs <- unique(c(pgs_white,pgs_black))

# SD of EA before and after this process goes from 1 to 1.01, implying that the HRS is slightly more diverse in EA than the population and that Z scoring using sampling weights doesn't really matter

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
  mod_pgs[[eth]] <- map(form_pgs, 
                               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                                     \(x) svyglm_quiet(x, r_eth))
  names(mod_pgs_edyrs[[eth]]) <- pgs

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                                    \(x) svyglm_quiet(x, r_parents_eth))

@@

    form_pgs_gkids <- map(pgs, 
                       \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))
  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
                                    \(x) svyglm_quiet(x, r_child_eth))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "edu13-17")) |>
    mutate(edu = str_remove(term, ".*:"))

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "income_medHigh")) |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "marriedMarried")) |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "agefbHigh")) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

@@

calc_eth_cor <- function (weights, data) {
  coef_pgs_white <- pgs |> 
        map(\(x) reformulate(c(x), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "white")) |> 
        map2_dbl(pgs, \(x, y) coef(x)[[y]])
  coef_pgs_black <- pgs |> 
        map(\(x) reformulate(c(x), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "black")) |> 
        map2_dbl(pgs, \(x, y) coef(x)[[y]])

@@

calc_cors_edu <- function (weights, data, var) {
  coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x), response = var)) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

  coef_pgs_edu <- pgs |> 
        map(\(x) reformulate(c(x), response = "raedyrs")) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

calc_eth_diff <- function (weights, data) {
  coef_pgs_white <- pgs |> 
        map(\(x) reformulate(c(x), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "white")) |> 
        map_dbl(\(x) coef(x)[[2]])
  coef_pgs_black <- pgs |> 
        map(\(x) reformulate(c(x), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "black")) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

f_resp <- map(pgs, \(x) reformulate(c(x), response = "raevbrn"))
f_sibs <- map(pgs, \(x) reformulate(c(x), response = "r10sibs"))

@@

f_resp <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
f_gkids <- map(pgs, \(x) reformulate(c(x), response = "rlrs_div_h13gkid"))

@@

  # "rlrs ~ pgs:term + term + pc1 + pc2 +..."
  forms <-  map(cross_terms, 
                \(x) reformulate(c(x, term), response = "rlrs"))
  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])

@@

```{r}
#| label: fig-deltas
#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
#|   polygenic scores 
#|   on RLRS among different social groups. A positive $\delta$ means that
#|   the coefficient was larger for the disadvantaged group, where the score
#|   was sign-flipped to correlate negatively with education. Horizontal lines
#|   show medians.

deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, 
               position = position_jitter(width = 0.1, seed = 12345)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    geom_text(aes(label = outliers), size = 2, colour = "black", 
              position = position_jitter(width = 0.1, seed = 12345), 
              hjust = -0.2) +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn"), 
                                                    response = "rlrs"))
mod_pgs_control_agefb <- map(form_pgs_control_agefb,
                             \(x) svyglm_quiet(x, rand, subset = 
                                                 ethnicity == "white"))
tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
                          list_rbind() |> 
                          filter(term %in% pgs)

@@

reg_simple <- lm(rlrs ~ EA ,rand$variables)
reg_pc <- lm(rlrs ~ EA + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
# beta = -0.027651
reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, "*edu") |>
    map(\(x) reformulate(c(x, "edu"), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
    map(\(x) reformulate(c(x, "income_med"), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_mar <- paste0(pgs, "*married") |>
    map(\(x) reformulate(c(x, "married"), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
    map(\(x) reformulate(c(x, "agefb"), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

@@

# some of the PGS have estimates rounded to zero and are thus useless
correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)
correct_scores <- correct_scores[,c("pgs",
                                    "chip_h2",
                                    "twin_h2",
                                    "R2",
                                    "Source", 
                                    "PoldermanICF_ICD10_subchapter",
                                    "Category")]

@@

correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
#correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28
pgs_correct <- correct_scores$pgs

####################################
# Running the regressions
####################################

form_pgs_correct <- map(pgs_correct, \(x) reformulate(c(x), response = "rlrs"))

corrected_effects <-  map(form_pgs_correct,  \(x) svyglm_quiet(x, rand)) |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |> 
    list_rbind() |>
    filter(term %in% pgs) |>
    rename(pgs = term) |>
    left_join(correct_scores)

corrected_effects_snp <- corrected_effects |>
    mutate(estimate = estimate*sqrt(chip_h2/R2),
           std.error = std.error*sqrt(chip_h2/R2),
           conf.low = conf.low*sqrt(chip_h2/R2),
           conf.high = conf.high*sqrt(chip_h2/R2),
           heritability = "Chip")

corrected_effects_twin <- corrected_effects |>
    mutate(estimate = estimate*sqrt(twin_h2/R2),
           std.error = std.error*sqrt(twin_h2/R2),
           conf.low = conf.low*sqrt(twin_h2/R2),
           conf.high = conf.high*sqrt(twin_h2/R2),
           heritability = "Twin")

corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)

@@

corrected_effects <- corrected_effects |> 
  rename(trait = pgs,
         TwinH2_Source = Source,
         Heritability = heritability)

corrected_effects |>
  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
  readr::write_csv("Corrected_effects.csv")

@@

corrected_effects |>
  # This removes ADHD due to the low power of the score. 
  # Note ADHD predict RLRS after bonferroni! 
  # It also removes NEB
  filter(R2 > 0.005) |> 
  mutate(
    trait = nice_names[trait],
    trait = fct_reorder(trait, estimate)
  ) |>
  ggplot(aes(y = trait, x = estimate, xmin = conf.low, xmax = conf.high,
           color = Heritability)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_pointrange(aes(linetype = Heritability == "Twin"),
                  position = position_dodge(0.2)) +
  scale_linetype(guide = "none") +
  coord_cartesian(xlim = c(-0.3, 0.3)) +
  labs(x = "Corrected coefficient on RLRS", y = "") +
  theme_minimal() +
  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
  theme(strip.text.y = element_text(angle = -90,face = "bold"))

@@

  fs_fert <- map(pgs, \(x) reformulate(c(x), response = "rlrs"))
  fs_phen <- map(pgs, \(x) reformulate(c(x), response = phenotype))

@@

```{r}
#| label: fig-true-effects-old
#| eval: false
#| fig-cap: |
#|   Estimated coefficients of true polygenic scores on RLRS calculated
#|   using estimates of chip- and twin-heritability and the most recent
#|   polygenic score.

# TODO: maybe just use the polygenic score name not the phenotype?
# TODO: why are some result bounds negative? e.g. ADHD? Presumably
#. because can have negative effect on RLRS?
# TODO: check the original phenotypes from which h2 ests were derived...
#. - is CAD the same as coronary heart disease (r10hearte)?
# TODO: why is ANXCC not predicting anxiety?
corrected_effects |> 
  mutate(
    Phenotype = nice_names[source.pgs],
    Heritability = target.pgs
  ) |> 
  filter(Heritability %in% c("Chip", "Twin"), ! source.pgs %in% old_pgs) |> 
  ggplot(aes(y = Phenotype, x = mean, xmin = conf.low, xmax = conf.high,
             color = Heritability)) +
    geom_vline(xintercept = 0, color = "grey60") +
    geom_pointrange(aes(linetype = Heritability == "Twin"),
                    position = position_dodge(0.2)) +
    scale_linetype(guide = "none") +
    coord_cartesian(xlim = c(- 0.3, 0.3)) +
    labs(x = "Corrected coefficient on RLRS", y = "") +
    theme_minimal()

@@

We also examined natural selection in different groups. Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time. The smaller black sample makes most tests inconclusive for this population: we can only say that the data do not reject a negative association between PGS correlations with RLRS and PGS correlations with education.

>>>>

---
title: "Natural Selection Across Three Generations of Americans"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for 
  natural selection: scores which correlate negatively (positively) with
  education are selected for (against). The theory predicts that
  selection coefficients should be stronger among low-income, 
  less educated, unmarried and younger parents, but these predictions are only
  partly borne out: coefficients are larger among low-income
  parents, but not among other groups. Using respondents' number of 
  siblings and grandchildren, we 
  examine selection across three generations, finding consistent results 
  over time. We also estimate effect sizes corrected for noise in the
  polygenic scores. Selection for some health traits is similar in magnitude
  to that for cognitive traits. 
date: "February 2024"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

@@

  form_pgs <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
  mod_pgs[[eth]] <- map(form_pgs, 
                               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_edyrs <- map(pgs, \(x) reformulate(c(x, pcs),
                                              response = "raedyrs"))
  mod_pgs_edyrs[[eth]] <- map(form_pgs_edyrs, 
                                     \(x) svyglm_quiet(x, r_eth))
  names(mod_pgs_edyrs[[eth]]) <- pgs

@@

  form_pgs_sibs <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "r10sibs"))
  mod_pgs_sibs[[eth]] <- map(form_pgs_sibs,
                                    \(x) svyglm_quiet(x, r_parents_eth))

@@

    form_pgs_gkids <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))
  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
                                    \(x) svyglm_quiet(x, r_child_eth))

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, ":edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_ed[[eth]] <-
    map(mod_pgs_x_ed[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_ed[[eth]]) <- pgs
  tidy_pgs_x_ed[[eth]] <-
    list_rbind(tidy_pgs_x_ed[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "edu13-17", pcs)) |>
    mutate(edu = str_remove(term, ".*:"))

  form_pgs_x_inc <- paste0(pgs, ":income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_inc[[eth]] <-
    map(mod_pgs_x_inc[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_inc[[eth]]) <- pgs
  tidy_pgs_x_inc[[eth]] <-
    list_rbind(tidy_pgs_x_inc[[eth]], names_to = "pgs") |>
    filter(!term %in% c("(Intercept)", "income_medHigh", pcs)) |>
    mutate(income_med = str_remove(term, ".*:income_med"))

  form_pgs_x_mar <- paste0(pgs, ":married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    ) 
}

@@

calc_eth_cor <- function (weights, data) {
  coef_pgs_white <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "white")) |> 
        map2_dbl(pgs, \(x, y) coef(x)[[y]])
  coef_pgs_black <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "black")) |> 
        map2_dbl(pgs, \(x, y) coef(x)[[y]])

@@

calc_cors_edu <- function (weights, data, var) {
  coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = var)) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

  coef_pgs_edu <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "raedyrs")) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

calc_eth_diff <- function (weights, data) {
  coef_pgs_white <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "white")) |> 
        map_dbl(\(x) coef(x)[[2]])
  coef_pgs_black <- pgs |> 
        map(\(x) reformulate(c(x, pcs), response = "rlrs")) |> 
        map(\(x) lm(x, data, weights = weights, 
                    subset = ethnicity == "black")) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
f_sibs <- map(pgs, \(x) reformulate(c(x, pcs), response = "r10sibs"))

@@

f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
f_gkids <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs_div_h13gkid"))

@@

  # "rlrs ~ pgs:term + term + pc1 + pc2 +..."
  forms <-  map(cross_terms, 
                \(x) reformulate(c(x, term, pcs), response = "rlrs"))
  mods <- map(forms, \(x) lm(x, data, weights = w))
  cfs <- map(mods, coef)
  cfs_low <- map_dbl(cfs, \(x) x[str_detect(names(x), low_group)])
  cfs_high <- map_dbl(cfs, \(x) x[str_detect(names(x), high_group)])

@@

```{r}
#| label: fig-deltas
#| fig-cap: Coefficients of interaction terms ($\delta$) for effects of
#|   polygenic scores 
#|   on RLRS among different social groups. A positive $\delta$ means that
#|   the selection coefficient was larger for the disadvantaged group, in 
#|   the direction associated with less education. Horizontal lines
#|   show median coefficients.

deltas |> 
  filter(
    group != "born"
  ) |> 
  mutate(
    med_d = median(delta),
    Group = case_match(group,
              "agefb"      ~ "AFB",
              "edu"        ~ "Education",
              "income_med" ~ "Income",
              "married"    ~ "Marriage"
            ),
    outliers = ifelse(delta > 0.05, nice_names[PGS], ""),
    .by = group
  ) |> 
  ggplot(aes(Group, delta, colour = Group)) + 
    geom_hline(yintercept = 0, color='grey70') + 
    geom_point(alpha = 0.7, 
               position = position_jitter(width = 0.1, seed = 12345)) +
    geom_spoke(aes(y = med_d), angle = 0,  radius = 0.3, colour = "grey40") +
    geom_spoke(aes(y = med_d), angle = pi, radius = 0.3, colour = "grey40") +
    geom_text(aes(label = outliers), size = 2, colour = "black", 
              position = position_jitter(width = 0.1, seed = 12345), 
              hjust = -0.15) +
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      legend.position = "none"
    )
```

@@

form_pgs_control_agefb <- map(pgs, \(x) reformulate(c(x, "agefbn", pcs), 
                                                    response = "rlrs"))
mod_pgs_control_agefb <- map(form_pgs_control_agefb,
                             \(x) svyglm_quiet(x, rand, subset = 
                                                 ethnicity == "white"))
tidy_pgs_control_agefb <- map(mod_pgs_control_agefb, broom::tidy) |> 
                          list_rbind() |> 
                          filter(term %in% pgs)

@@

reg_simple <- lm(rlrs ~ scale(EA) ,rand$variables)
reg_pc <- lm(rlrs ~ scale(EA) + PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,rand$variables)
# beta = -0.027651
reg_res_EA <- lm(`scale(EA)` ~ PC1_5A +PC1_5B +PC1_5C +PC1_5D +PC1_5E + 
             PC6_10A +PC6_10B +PC6_10C +PC6_10D +PC6_10E ,reg_pc$model)

@@

  # the colon interaction estimates the effect separately within each group
  form_pgs_x_ed <- paste0(pgs, "*edu") |>
    map(\(x) reformulate(c(x, "edu", pcs), response = "rlrs"))
  mod_pgs_x_ed[[eth]]  <-
    map(form_pgs_x_ed, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_inc <- paste0(pgs, "*income_med") |>
    map(\(x) reformulate(c(x, "income_med", pcs), response = "rlrs"))
  mod_pgs_x_inc[[eth]]  <-
    map(form_pgs_x_inc, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_mar <- paste0(pgs, "*married") |>
    map(\(x) reformulate(c(x, "married", pcs), response = "rlrs"))
  mod_pgs_x_mar[[eth]]  <-
    map(form_pgs_x_mar, 
               \(x) svyglm_quiet(x, r_eth))

@@

  form_pgs_x_agefb <- paste0(pgs, "*agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "rlrs"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
               \(x) svyglm_quiet(x, r_eth))

@@

# some of the PGS have estimates rounded to zero and are thus useless
correct_scores <- filter(correct_scores, chip_h2 > 0, R2 > 0)

@@

correct_scores$twin_h2[correct_scores$pgs == "AFB"] <- 0.15
#correct_scores$twin_h2[correct_scores$pgs == "NEBwomen"] <- 0.28

####################################
# Calculating error corrected estimates
####################################

run_pgic_once <- function (pgs, r2, h2, h2_type) {
  pgic_command <- glue(
     'python3 pgi_correct/pgic.py --reg-data-file pgi_correct_data.txt', 
     '--outcome "rlrs" --pgi-var {pgs}',
     '--covariates "PC1_5A" "PC1_5B" "PC1_5C" "PC1_5D" "PC1_5E" "PC6_10A"',
     '"PC6_10B" "PC6_10C" "PC6_10D" "PC6_10E" --R2 {r2} --h2 {h2}',
     '--weights "weight"', 
     .sep = " ")
  system(pgic_command)

  # This removes "junk" rows in the way of our regression betas
  system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
  pgic_results <- readr::read_table("reg_results", col_names = TRUE,
                                    col_types = "cdddd")
  file.remove("reg_results")

  tibble(
    trait = pgs,
    beta  = pgic_results[[1, 4]],
    SE    = pgic_results[[1, 5]],
    Heritability = h2_type
  )
}

run_pgic_data <- function (score_data, h2_column, h2_type) {
  results <- map(1:nrow(score_data), 
    \(i) run_pgic_once(pgs = score_data$pgs[i], 
                       r2 = score_data$R2[i],
                       h2 = score_data[i, h2_column, drop = TRUE],
                       h2_type = h2_type)
  )

  list_rbind(results)
}

corrected_effects_snp <- run_pgic_data(correct_scores, h2_column = "chip_h2",
                                       h2_type = "Chip")
corrected_effects_twin <- run_pgic_data(correct_scores, h2_column = "twin_h2",
                                        h2_type = "Twin")

corrected_effects <- rbind(corrected_effects_snp, corrected_effects_twin)
corrected_effects$conf.low <- corrected_effects$beta - 1.96 *
                              corrected_effects$SE
corrected_effects$conf.high <- corrected_effects$beta + 1.96 *
                               corrected_effects$SE

unlink("pgi_correct_data.txt")

corrected_effects <- left_join(corrected_effects, correct_scores,
                               by = join_by(trait == Abr),
                               suffix = c("", ".y"))

# The odd looking results are for traits with large missing heritability e.g. Asthma. My guess is that the rare variants act somewhat differently as DHJ as argued, meaning traits with large missing heritability show nonsense twin heritability estimates.

@@

# Cleaning and saving corrected effects so people can have the exact figures
corrected_effects$TwinH2_Source <- corrected_effects$Source
corrected_effects |> 
  select(trait, beta, SE, Heritability, chip_h2, twin_h2, R2, 
           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
  readr::write_csv("Corrected_effects.csv")

@@

# TE Note that under Fisher's fundamental theorem of natural selection, the selection
# for fitness should be equal to it's heritability... But in HRS the
# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
# is much lower than what the heritability of NEB should be.
# The low R2 of NEB causes it to be removed, but the genetics for NEB 
# could just be a little different in HRS versus other samples!
# NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]

corrected_effects |>
  # This removes ADHD due to the low power of the score. 
  # Note ADHD predict RLRS after bonferroni! 
  # It also removes NEB
  filter(R2 > 0.005) |> 
  mutate(
    trait = nice_names[trait],
    trait = fct_reorder(trait, beta)
  ) |>
  ggplot(aes(y = trait, x = beta, xmin = conf.low, xmax = conf.high,
           color = Heritability)) +
  geom_vline(xintercept = 0, color = "grey60") +
  geom_pointrange(aes(linetype = Heritability == "Twin"),
                  position = position_dodge(0.2)) +
  scale_linetype(guide = "none") +
  coord_cartesian(xlim = c(-0.3, 0.3)) +
  labs(x = "Corrected coefficient on RLRS", y = "") +
  theme_minimal() +
  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
  theme(strip.text.y = element_text(angle = -90,face = "bold"))

@@

  fs_fert <- map(pgs, \(x) reformulate(c(x, pcs), response = "rlrs"))
  fs_phen <- map(pgs, \(x) reformulate(c(x, pcs), response = phenotype))

@@

@@

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time.

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is similar to previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.
